---
title: "Are Harris County Drivers the Worst in Texas?"
subtitle: "An Editable Beamer Presentation"
author: "Fredy Guevara"
date: "`r Sys.Date()`"
classoption: aspectratio=169
output:
  beamer_presentation:
    theme: Madrid
    colortheme: default
    fonttheme: structurebold
    keep_tex: true
header-includes:
  - \setbeamertemplate{footline}{}
  - \setbeamertemplate{navigation symbols}{}
  - \definecolor{RoadYellow}{RGB}{242,204,0}
  - \definecolor{RoadBlack}{RGB}{20,20,20}
  - \definecolor{RoadGray}{RGB}{245,245,245}
  - \setbeamercolor{palette primary}{bg=RoadBlack, fg=RoadYellow}
  - \setbeamercolor{palette secondary}{bg=RoadBlack, fg=white}
  - \setbeamercolor{palette tertiary}{bg=RoadYellow, fg=RoadBlack}
  - \setbeamercolor{palette quaternary}{bg=RoadBlack, fg=white}
  - \setbeamercolor{structure}{fg=RoadYellow}
  - \setbeamercolor{title}{fg=RoadYellow}
  - \setbeamercolor{frametitle}{bg=RoadBlack, fg=RoadYellow}
  - \setbeamercolor{normal text}{fg=black, bg=white}
  - \setbeamercolor{block title}{bg=RoadBlack, fg=RoadYellow}
  - \setbeamercolor{block body}{bg=RoadGray, fg=black}
  - \setbeamercolor{item}{fg=RoadYellow}
  - \setbeamertemplate{footline}{}
  - \setbeamertemplate{navigation symbols}{}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 4.8,
  fig.align = "center"
)

library(tidyverse)

# ---- Data import ----
crash <- read.csv("Crash_Data_2024.csv")

# ---- Data cleaning ----
crash <- crash %>%
  mutate(
    Total_Crashes = as.numeric(gsub(",", "", Total_Crashes)),
    Population = as.numeric(gsub(",", "", Population)),
    Fatal_Crashes = as.numeric(gsub(",", "", Fatal_Crashes)),
    Speed_Crashes = as.numeric(gsub(",", "", Speed_Crashes)),
    DUI_Crashes = as.numeric(gsub(",", "", DUI_Crashes)),
    Distracted_Crashes = as.numeric(gsub(",", "", Distracted_Crashes)),
    Serious_Crashes = as.numeric(gsub(",", "", Serious_Crashes)),
    Minor_Crashes = as.numeric(gsub(",", "", Minor_Crashes))
  ) %>%
  mutate(
    Crash_Rate = (Total_Crashes / Population) * 100000,
    Fatal_Rate = (Fatal_Crashes / Population) * 100000,
    SpeedPercent = Speed_Crashes / Total_Crashes,
    DUIPercent = DUI_Crashes / Total_Crashes,
    DistractedPercent = Distracted_Crashes / Total_Crashes,
    SeverityPercent = (Fatal_Crashes + Serious_Crashes) / Total_Crashes
  )

# ---- Ranked tables ----
rank_crash <- crash %>%
  arrange(desc(Crash_Rate)) %>%
  select(County, Crash_Rate, Total_Crashes, Population)

rank_fatal <- crash %>%
  arrange(desc(Fatal_Rate)) %>%
  select(County, Fatal_Rate, Fatal_Crashes, Population)

risk_table <- crash %>%
  select(County, SpeedPercent, DUIPercent, DistractedPercent, SeverityPercent) %>%
  arrange(desc(SpeedPercent))

# ---- Confidence intervals ----
make_ci <- function(df, county_name) {
  row <- df %>% filter(County == county_name)
  p_hat <- row$Total_Crashes / row$Population
  n <- row$Population
  z <- 1.96
  se <- sqrt((p_hat * (1 - p_hat)) / n)
  tibble(
    County = county_name,
    Lower_CI = (p_hat - z * se) * 100000,
    Upper_CI = (p_hat + z * se) * 100000,
    Crash_Rate = p_hat * 100000
  )
}

ci_compare <- bind_rows(
  make_ci(crash, "Harris"),
  make_ci(crash, "Montgomery")
)

# ---- Regression ----
model <- lm(Crash_Rate ~ SpeedPercent + DUIPercent + DistractedPercent, data = crash)
model_coef <- broom::tidy(model)
model_fit <- broom::glance(model)

# ---- Key result values for slides ----
top_crash <- crash %>%
  arrange(desc(Crash_Rate)) %>%
  select(County, Crash_Rate)

harris_rate <- crash %>% filter(County == "Harris") %>% pull(Crash_Rate)
bexar_rate <- crash %>% filter(County == "Bexar") %>% pull(Crash_Rate)
travis_rate <- crash %>% filter(County == "Travis") %>% pull(Crash_Rate)
lipscomb_rate <- crash %>% filter(County == "Lipscomb") %>% pull(Crash_Rate)

r2_val <- model_fit$r.squared

ci_all <- crash %>%
  mutate(
    p_hat = Total_Crashes / Population,
    n = Population,
    z = 1.96,
    se = sqrt((p_hat * (1 - p_hat)) / n),
    Lower_CI = (p_hat - z * se) * 100000,
    Upper_CI = (p_hat + z * se) * 100000,
    Crash_Rate = p_hat * 100000
  ) %>%
  select(County, Crash_Rate, Lower_CI, Upper_CI)
```

## 

### Are Harris County Drivers the Worst in Texas?

Senior project comparing selected Texas counties using crash rates, confidence intervals, and regression.

\vspace{0.5cm}

Selected counties: Harris, Travis, Bexar, Montgomery, Hays, Medina, Hidalgo, Hale, and Lipscomb.

## Research Question

### Main Question

Is Harris County the county with the worst drivers among the counties in this sample?

### Working Idea

This project treats “worst drivers” as a question of:

- crash frequency relative to population
- fatal crash risk
- risky driving patterns such as speeding, DUI, and distraction

## Why Total Crashes Are Not Enough

A large county will usually have more crashes simply because it has more people and more traffic.

That means total crashes alone can be misleading.

### Better comparison measures

- crash rate per 100,000 people
- fatal crash rate per 100,000 people
- proportions involving speeding, DUI, and distracted driving



## Counties in the Study

```{r counties-slide}
knitr::kable(
  tibble(County = unique(crash$County)),
  booktabs = TRUE,
  align = "l",
  caption = "Counties included in the analysis"
)
```



## Creating New Variables

To compare counties fairly, several derived variables were created.

\[
\text{Crash Rate} = \frac{\text{Total Crashes}}{\text{Population}} \times 100{,}000
\]

\[
\text{Fatal Crash Rate} = \frac{\text{Fatal Crashes}}{\text{Population}} \times 100{,}000
\]

\[
\text{Speed Percent} = \frac{\text{Speed Crashes}}{\text{Total Crashes}}
\]

\[
\text{DUI Percent} = \frac{\text{DUI Crashes}}{\text{Total Crashes}}
\]

\[
\text{Distracted Percent} = \frac{\text{Distracted Crashes}}{\text{Total Crashes}}
\]




## Summary of Derived Variables

```{r derived-table}
knitr::kable(
  crash %>%
    select(County, Crash_Rate, Fatal_Rate, SpeedPercent, DUIPercent, DistractedPercent),
  digits = 3,
  caption = "Derived Variables"
)
```


## Ranking Counties by Crash Rate

```{r crash-rank-table}
knitr::kable(
  rank_crash %>% mutate(Crash_Rate = round(Crash_Rate, 2)),
  booktabs = TRUE,
  caption = "Counties ranked by crash rate per 100,000"
)
```

## Crash Rate per 100,000 People

- Harris County: `r round(harris_rate, 1)` (highest)
- Bexar County: `r round(bexar_rate, 1)` (very close)
- Travis County: `r round(travis_rate, 1)` (lowest among large counties)
- Lipscomb County: `r round(lipscomb_rate, 1)` (lowest overall)

Harris has the highest crash rate, but only slightly higher than Bexar.

## Crash Rate by County

```{r, fig.width=5, fig.height=3}
ggplot(crash, aes(x = reorder(County, Crash_Rate), y = Crash_Rate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Crash Rate per 100,000 by County",
    x = "County",
    y = "Crash Rate"
  ) +
  theme_minimal(base_size = 16)

```

## Does Harris Have the Worst Drivers?

- Harris County has the highest crash rate
- However:
  - difference from Bexar is small
  - not dramatically worse than other large counties

### Conclusion

Houston is not overwhelmingly worse, just slightly higher.

## Ranking Counties by Fatal Crash Rate

```{r fatal-rank-table}
knitr::kable(
  rank_fatal %>% mutate(Fatal_Rate = round(Fatal_Rate, 2)),
  booktabs = TRUE,
  caption = "Counties ranked by fatal crash rate per 100,000"
)
```

### Brief interpretation

Fatal crash rate focuses on severity rather than frequency, so it can highlight counties where crashes are less common but more dangerous.

## Risk Factor Table

```{r risk-factor-table}
knitr::kable(
  risk_table %>% mutate(across(-County, ~ round(.x, 4))),
  booktabs = TRUE,
  caption = "Risk-factor proportions by county"
)
```

## Driving Behavior Across Counties

- Speed-related crashes:
  - moderate across most counties
- DUI crashes:
  - relatively low percentages
- Distracted driving:
  - high in multiple counties, especially urban counties

### Insight

Risk behaviors vary, but no single factor dominates.

## Speeding vs Crash Rate

```{r, fig.width=5, fig.height=3}
ggplot(crash, aes(x = SpeedPercent, y = Crash_Rate, label = County)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(nudge_y = 35, size = 3) +
  labs(
    title = "Speeding vs Crash Rate",
    x = "Proportion of Crashes Involving Speed",
    y = "Crash Rate per 100,000"
  ) +
  theme_minimal(base_size = 16)
```

### Brief interpretation

An upward trend would suggest that counties with more speed-related crashes also tend to have higher crash rates.

## Distracted Driving vs Crash Rate

```{r, fig.width=5, fig.height=3}
ggplot(crash, aes(x = DistractedPercent, y = Crash_Rate, label = County)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(nudge_y = 35, size = 3) +
  labs(
    title = "Distracted Driving vs Crash Rate",
    x = "Proportion of Crashes Involving Distracted Driving",
    y = "Crash Rate per 100,000"
  ) +
  theme_minimal(base_size = 16)
```

### Brief interpretation

This plot checks whether distracted driving is associated with higher crash rates across the sample counties.

## Confidence Intervals for County Crash Rates

To estimate the true crash rate for each county, I used a 95\% confidence interval.

\[
\hat{p}=\frac{\text{Total Crashes}}{\text{Population}}
\]

\[
\hat{p}\pm 1.96\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
\]

where:

- \(\hat{p}\) = observed crash proportion
- \(n\) = county population
- \(1.96\) gives a 95\% confidence level

### Purpose

This helps show whether differences in county crash rates are likely to be meaningful or just due to random variation. 

## 95\% Confidence Intervals by County

```{r ci-all-table, results='asis'}
knitr::kable(
  ci_all %>% mutate(across(-County, ~ round(.x, 2))),
  format = "latex",
  booktabs = TRUE,
  caption = "Crash rates and 95% confidence intervals per 100,000 residents"
)
```



## Crash Rates with 95\% Confidence Intervals

```{r ci-all-plot, fig.width=6.8, fig.height=4.2}
ggplot(ci_all, aes(x = reorder(County, Crash_Rate), y = Crash_Rate)) +
  geom_point(size = 2.8) +
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), width = 0.2) +
  coord_flip() +
  labs(
    title = "Crash Rates with 95% Confidence Intervals",
    x = "County",
    y = "Crash Rate per 100,000"
  ) +
  theme_minimal(base_size = 15)
```
## Interpreting the Confidence Intervals

- If two counties have intervals that do not overlap, that suggests stronger evidence that their crash rates are different
- If two counties have intervals that overlap a lot, then the difference may not be statistically meaningful

### Harris vs Montgomery

- Harris County has a higher observed crash rate than Montgomery County
- But if their intervals overlap, then the data does not clearly show that Harris is worse
- This matters because neighboring counties may share highways, commuter traffic, and similar driving environments

### Main Point

Confidence intervals help move the project beyond simple ranking and toward statistical comparison.

## Regression Model

\[
\text{Crash Rate}=\beta_0+\beta_1(\text{SpeedPercent})+\beta_2(\text{DUIPercent})+\beta_3(\text{DistractedPercent})+\varepsilon
\]

- Goal: identify which behaviors increase crash rates

## Are the results significant?

```{r regression-table}
knitr::kable(
  model_coef %>% mutate(across(where(is.numeric), ~ round(.x, 4))),
  booktabs = TRUE,
  caption = "Regression coefficients"
)
```

- All p-values > 0.05
- None of the variables are statistically significant

### Interpretation

We cannot confidently say any factor causes higher crash rates.

## Regression Findings

```{r model-fit}
knitr::kable(
  model_fit %>%
    select(r.squared, adj.r.squared, sigma, statistic, p.value) %>%
    mutate(across(everything(), ~ round(.x, 4))),
  booktabs = TRUE,
  caption = "Regression model fit"
)
```

- \(R^2 = `r round(r2_val, 2)`\)
- Model explains only about `r round(r2_val*100, 0)`\% of the variation

### Coefficients

- Speed → positive effect
- Distracted → positive effect
- DUI → unclear or negative effect

## Interpretation of Results

Crash rates are not fully explained by:

- Speed
- DUI
- Distracted driving

### Likely Explanation

Other factors may matter more:

- Traffic density
- Urban congestion
- Road design

## Limitations

- Small sample size (\(n = 9\) counties)
- Limited variables
- Regression results are not statistically significant

### Implication

The results suggest trends, but not definitive causation.

## Final Conclusion

- Harris County has the highest crash rate
- But:
  - only slightly higher than Bexar
  - not strongly explained by driver behavior

### Final Answer

Harris does not clearly have the “worst drivers.”




