---
title: "Are Harris County Drivers the Worst in Texas?"
subtitle: "An Editable Beamer Presentation"
author: "Fredy Guevara"
date: "`r Sys.Date()`"
classoption: aspectratio=169
output:
  beamer_presentation:
    theme: Madrid
    colortheme: dolphin
    fonttheme: structurebold
    keep_tex: true
header-includes:
  - \setbeamertemplate{footline}{}
  - \setbeamertemplate{navigation symbols}{}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 4.8,
  fig.align = "center"
)

library(tidyverse)

# ---- Data import ----
crash <- read.csv("Crash_Data_2024.csv")

# ---- Data cleaning ----
crash <- crash %>%
  mutate(
    Total_Crashes = as.numeric(gsub(",", "", Total_Crashes)),
    Population = as.numeric(gsub(",", "", Population)),
    Fatal_Crashes = as.numeric(gsub(",", "", Fatal_Crashes)),
    Speed_Crashes = as.numeric(gsub(",", "", Speed_Crashes)),
    DUI_Crashes = as.numeric(gsub(",", "", DUI_Crashes)),
    Distracted_Crashes = as.numeric(gsub(",", "", Distracted_Crashes)),
    Serious_Crashes = as.numeric(gsub(",", "", Serious_Crashes)),
    Minor_Crashes = as.numeric(gsub(",", "", Minor_Crashes))
  ) %>%
  mutate(
    Crash_Rate = (Total_Crashes / Population) * 100000,
    Fatal_Rate = (Fatal_Crashes / Population) * 100000,
    SpeedPercent = Speed_Crashes / Total_Crashes,
    DUIPercent = DUI_Crashes / Total_Crashes,
    DistractedPercent = Distracted_Crashes / Total_Crashes,
    SeverityPercent = (Fatal_Crashes + Serious_Crashes) / Total_Crashes
  )

# ---- Ranked tables ----
rank_crash <- crash %>%
  arrange(desc(Crash_Rate)) %>%
  select(County, Crash_Rate, Total_Crashes, Population)

rank_fatal <- crash %>%
  arrange(desc(Fatal_Rate)) %>%
  select(County, Fatal_Rate, Fatal_Crashes, Population)

risk_table <- crash %>%
  select(County, SpeedPercent, DUIPercent, DistractedPercent, SeverityPercent) %>%
  arrange(desc(SpeedPercent))

# ---- Confidence intervals ----
make_ci <- function(df, county_name) {
  row <- df %>% filter(County == county_name)
  p_hat <- row$Total_Crashes / row$Population
  n <- row$Population
  z <- 1.96
  se <- sqrt((p_hat * (1 - p_hat)) / n)
  tibble(
    County = county_name,
    Lower_CI = (p_hat - z * se) * 100000,
    Upper_CI = (p_hat + z * se) * 100000,
    Crash_Rate = p_hat * 100000
  )
}

ci_compare <- bind_rows(
  make_ci(crash, "Harris"),
  make_ci(crash, "Montgomery")
)

# ---- Regression ----
model <- lm(Crash_Rate ~ SpeedPercent + DUIPercent + DistractedPercent, data = crash)
model_coef <- broom::tidy(model)
model_fit <- broom::glance(model)
```

## 

### Are Harris County Drivers the Worst in Texas?

Senior project comparing selected Texas counties using crash rates, confidence intervals, and regression.

\vspace{0.5cm}

Selected counties: Harris, Travis, Bexar, Montgomery, Hays, Medina, Hidalgo, Hale, and Lipscomb.

## Research Question

### Main Question

Is Harris County the county with the worst drivers among the counties in this sample?

### Working Idea

This project treats “worst drivers” as a question of:

- crash frequency relative to population
- fatal crash risk
- risky driving patterns such as speeding, DUI, and distraction

## Why Total Crashes Are Not Enough

A large county will usually have more crashes simply because it has more people and more traffic.

That means total crashes alone can be misleading.

### Better comparison measures

- crash rate per 100,000 people
- fatal crash rate per 100,000 people
- proportions involving speeding, DUI, and distracted driving



## Counties in the Study

```{r counties-slide}
knitr::kable(
  tibble(County = unique(crash$County)),
  booktabs = TRUE,
  align = "l",
  caption = "Counties included in the analysis"
)
```



## Creating New Variables

To compare counties fairly, several derived variables were created.

\[
\text{Crash Rate} = \frac{\text{Total Crashes}}{\text{Population}} \times 100{,}000
\]

\[
\text{Fatal Crash Rate} = \frac{\text{Fatal Crashes}}{\text{Population}} \times 100{,}000
\]

\[
\text{Speed Percent} = \frac{\text{Speed Crashes}}{\text{Total Crashes}}
\]

\[
\text{DUI Percent} = \frac{\text{DUI Crashes}}{\text{Total Crashes}}
\]

\[
\text{Distracted Percent} = \frac{\text{Distracted Crashes}}{\text{Total Crashes}}
\]

## Severity Measure

To capture how dangerous crashes are, I also used:

\[
\text{Severity Percent} = \frac{\text{Fatal Crashes} + \text{Serious Crashes}}{\text{Total Crashes}}
\]

This measure helps distinguish between counties that have many crashes and counties that have more severe crashes.

## Summary of Derived Variables

```{r derived-table}
knitr::kable(
  crash %>%
    select(County, Crash_Rate, Fatal_Rate, SpeedPercent, DUIPercent, DistractedPercent, SeverityPercent) %>%
    mutate(across(where(is.numeric), ~ round(.x, 4))) %>%
    arrange(desc(Crash_Rate)),
  booktabs = TRUE,
  caption = "Derived variables used in the analysis"
)
```

## Ranking Counties by Crash Rate

```{r crash-rank-table}
knitr::kable(
  rank_crash %>% mutate(Crash_Rate = round(Crash_Rate, 2)),
  booktabs = TRUE,
  caption = "Counties ranked by crash rate per 100,000"
)
```

### Brief interpretation

Ranking by crash rate is more meaningful than ranking by total crashes because it adjusts for county size.

## Crash Rate by County

```{r crash-rate-plot}
ggplot(crash, aes(x = reorder(County, Crash_Rate), y = Crash_Rate)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Crash Rate per 100,000 by County",
    x = "County",
    y = "Crash Rate"
  ) +
  theme_minimal(base_size = 16)
```

### Brief interpretation

If Harris County is near the top of this chart, that supports the claim that Houston-area drivers face a higher crash risk. If several counties are close, then Harris may not be uniquely worst.

## Ranking Counties by Fatal Crash Rate

```{r fatal-rank-table}
knitr::kable(
  rank_fatal %>% mutate(Fatal_Rate = round(Fatal_Rate, 2)),
  booktabs = TRUE,
  caption = "Counties ranked by fatal crash rate per 100,000"
)
```

### Brief interpretation

Fatal crash rate focuses on severity rather than frequency, so it can highlight counties where crashes are less common but more dangerous.

## Risk Factor Table

```{r risk-factor-table}
knitr::kable(
  risk_table %>% mutate(across(-County, ~ round(.x, 4))),
  booktabs = TRUE,
  caption = "Risk-factor proportions by county"
)
```

### Brief interpretation

These percentages help explain why crash rates differ across counties. High values suggest that risky behavior may contribute to county-level differences.

## Speeding vs Crash Rate

```{r speed-vs-crash}
ggplot(crash, aes(x = SpeedPercent, y = Crash_Rate, label = County)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(nudge_y = 35, size = 3) +
  labs(
    title = "Speeding vs Crash Rate",
    x = "Proportion of Crashes Involving Speed",
    y = "Crash Rate per 100,000"
  ) +
  theme_minimal(base_size = 16)
```

### Brief interpretation

An upward trend would suggest that counties with more speed-related crashes also tend to have higher crash rates.

## Distracted Driving vs Crash Rate

```{r distracted-vs-crash}
ggplot(crash, aes(x = DistractedPercent, y = Crash_Rate, label = County)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_text(nudge_y = 35, size = 3) +
  labs(
    title = "Distracted Driving vs Crash Rate",
    x = "Proportion of Crashes Involving Distracted Driving",
    y = "Crash Rate per 100,000"
  ) +
  theme_minimal(base_size = 16)
```

### Brief interpretation

This plot checks whether distracted driving is associated with higher crash rates across the sample counties.

## Confidence Interval Setup

To estimate the true county crash proportion, I used:

\[
\hat{p} = \frac{\text{Total Crashes}}{\text{Population}}
\]

\[
\hat{p} \pm z \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
\]

where \(z = 1.96\) for a 95\% confidence level.

## Confidence Intervals: Harris vs Montgomery

```{r ci-table}
knitr::kable(
  ci_compare %>% mutate(across(-County, ~ round(.x, 2))),
  booktabs = TRUE,
  caption = "95% confidence intervals for crash rate per 100,000"
)
```

### Brief interpretation

If the two intervals overlap a lot, then the difference between Harris and Montgomery may not be statistically meaningful.

## Regression Model

The regression model used was:

\[
\text{Crash Rate} = \beta_0 + \beta_1(\text{SpeedPercent}) + \beta_2(\text{DUIPercent}) + \beta_3(\text{DistractedPercent}) + \varepsilon
\]

This model tests whether crash rate is associated with speeding, DUI, and distracted-driving shares.

## Regression Output

```{r regression-table}
knitr::kable(
  model_coef %>% mutate(across(where(is.numeric), ~ round(.x, 4))),
  booktabs = TRUE,
  caption = "Regression coefficients"
)
```

## Model Fit

```{r model-fit}
knitr::kable(
  model_fit %>%
    select(r.squared, adj.r.squared, sigma, statistic, p.value) %>%
    mutate(across(everything(), ~ round(.x, 4))),
  booktabs = TRUE,
  caption = "Regression model fit"
)
```

### Brief interpretation

The coefficients show the direction and size of each relationship. Smaller p-values suggest stronger evidence that a variable matters. The \(R^2\) value shows how much of the variation in crash rates is explained by the model.

## Main Takeaways

- Total crashes are not enough to identify the “worst drivers.”
- Crash rate and fatal crash rate provide fairer county comparisons.
- Risk-factor percentages help explain why counties differ.
- Harris County should be judged by rate-based measures, not just size.

## Limitations

- The sample includes selected counties rather than all Texas counties.
- County-level analysis cannot identify individual driver behavior.
- Population-based crash rates are useful, but vehicle miles traveled would be an even stronger exposure measure.
- Small counties may produce unstable percentages because of low counts.

## Conclusion

This project uses rates, confidence intervals, and regression to study whether Harris County stands out as having the worst drivers among the selected counties.

The final answer depends on whether Harris ranks highest after adjusting for population and whether its estimated crash rate is clearly different from comparable counties.




